Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-05-26 papers/wdmerger_I/paper.tex
remove extra )
Add a directory for paper II
We were writing down the wrong formalism for the rotation--energy coupling; it is a static force and so it does not make sense to use the same approach we use for gravity, which dynamically responds to the movement of fluid on the grid.

We propose a more accurate coupling that directly solves this implicit system of coupled 
equations. We denote by $(\widetilde{\rho \mathbf{u}})$ the value of the momentum after 
updating it with the centrifugal force, and the time-level $n$ Coriolis force. The remaining 
update for the time-level $n+1$ Coriolis force then appears as:
\begin{equation}
  (\rho \mathbf{u})^{n+1} = (\widetilde{\rho\mathbf{u}}) + \frac{\Delta t}{2} \left(-2\, {\bm\omega} \times (\rho\mathbf{u})^{n+1}\right)
\end{equation}
To proceed further, we assume that the rotation is about the $z$ axis with frequency $\omega$. 
Then there is no update to the $z$-momentum, and the other equations are:
\begin{align}
  (\rho u)^{n+1} &= (\widetilde{\rho u}) + \omega \Delta t (\rho v)^{n+1}) \\
  (\rho v)^{n+1} &= (\widetilde{\rho v}) - \omega \Delta t (\rho u)^{n+1})
\end{align}
We can directly solve this coupled system:
\begin{align}
  (\rho u)^{n+1} &= \frac{ (\widetilde{\rho u}) + \omega \Delta t (\widetilde{\rho v})}{1 + \omega^2 \Delta t^2} \\
  (\rho v)^{n+1} &= \frac{ (\widetilde{\rho v}) - \omega \Delta t (\widetilde{\rho u})}{1 + \omega^2 \Delta t^2}
\end{align}
We use this form of the momentum update in CASTRO. This improvement is small
but increases the accuracy of our rotating white dwarf systems over long time-scales.

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, and so
 the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}_R\, dV,
\end{equation}
with $\mathbf{f}_R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}_R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi_R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
Therefore we can retrace the steps of Section \ref{sec:gravity_hydro_coupling} to obtain
an update to $(\rho E)$ that conserves the total energy in the presence of rotation,
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi_R.
\end{equation}
In one dimension, this would be
\begin{align}
  \left.\Delta(\rho E_i)\right|_{\text{rot}} &= \frac{1}{2} \Delta m_{i-1/2} (\Phi_{R_{i}} - \Phi_{R_{i-1}}) \notag \\
  &+ \frac{1}{2} \Delta m_{i+1/2} (\Phi_{R_{i+1}} - \Phi_{R_{i}}).
\end{align}
As with gravity, in three dimensions we include the mass fluxes through all six cell faces.

>>>>

We propose a more accurate coupling that directly solves this implicit system of coupled 
equations. We denote by $(\widetilde{\rho \mathbf{u}})$ the value of the momentum after 
updating it with the centrifugal force, and the time-level $n$ Coriolis force. The remaining 
update for the time-level $n+1$ Coriolis force then appears as:
\begin{equation}
  (\rho \mathbf{u})^{n+1} = (\widetilde{\rho\mathbf{u}}) + \frac{\Delta t}{2} \left(-2\, {\bm\omega} \times (\rho\mathbf{u})^{n+1}\right)
\end{equation}
To proceed further, we assume that the rotation is about the $z$ axis with frequency $\omega$. 
Then there is no update to the $z$-momentum, and the other equations are:
\begin{align}
  (\rho u)^{n+1} &= (\widetilde{\rho u}) + \omega \Delta t (\rho v)^{n+1} \\
  (\rho v)^{n+1} &= (\widetilde{\rho v}) - \omega \Delta t (\rho u)^{n+1}
\end{align} 
We can directly solve this coupled system:
\begin{align}
  (\rho u)^{n+1} &= \frac{ (\widetilde{\rho u}) + \omega \Delta t (\widetilde{\rho v})}{1 + \omega^2 \Delta t^2} \\
  (\rho v)^{n+1} &= \frac{ (\widetilde{\rho v}) - \omega \Delta t (\widetilde{\rho u})}{1 + \omega^2 \Delta t^2}
\end{align}
We use this form of the momentum update in CASTRO. This improvement is small
but increases the accuracy of our rotating white dwarf systems over long time-scales.

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, and so
 the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}_R\, dV,
\end{equation}
with $\mathbf{f}_R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical in form to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}_R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi_R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
So in the presence of rotation the total energy takes the form:
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi_R.
\end{equation}
Since the rotation field $\Phi_R$ is static in time and does not change in response 
to moving fluid, we do not need the formalism of Section \ref{sec:gravity_hydro_coupling}. 
Instead, the time rate of change in the energy can be written as the flux of the quantity 
$\rho \Phi_R$. In one dimension, this would be
\begin{align}
  \left.\Delta(\rho E_i)\right|_{\text{rot}} &= \frac{1}{2} \Delta m_{i-1/2} (\Phi_{R_{i}} + \Phi_{R_{i-1}}) \notag \\
  &+ \frac{1}{2} \Delta m_{i+1/2} (\Phi_{R_{i+1}} + \Phi_{R_{i}}).
\end{align}
As with gravity, in three dimensions we include the mass fluxes through all six cell faces.
Note that whereas the gravity force explicitly changes the gas energy $E_g$, the rotation force 
is manifestly conservative with respect to $E_g$ (except at domain boundaries), and only 
redistributes existing quantities of $E_g$ between zones. This asymmetry between the gravity 
and rotation forces may not be aesthetically pleasing but is necessary to avoid the problems 
mentioned earlier that come from explicitly evolving the equation for $E_\text{tot}$.
